\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{graphicx}

% Figures drawn as vector diagrams (no external binaries required)
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.multipart}

\title{Clinical Condition Extraction from Longitudinal Patient Notes}
\author{NLP Assignment Submission}
\date{\today}

\begin{document}
\maketitle

\section{Overview}

We are given longitudinal clinical notes per patient (2--13 notes), ordered chronologically by filename (\texttt{text\_0.md} earliest). The task is to extract a structured condition summary per patient and write one JSON file \texttt{patient\_XX.json} following the schema in \texttt{clinical\_nlp\_assignment/problem\_statement.md}. Each condition must:
\begin{itemize}[leftmargin=1.2em]
  \item map to exactly one taxonomy slot (\texttt{category.subcategory}) from \texttt{taxonomy.json};
  \item include a \texttt{status} in \{\texttt{active}, \texttt{resolved}, \texttt{suspected}\} reflecting the latest note where it appears;
  \item include an \texttt{onset} date using the specified priority rules (or \texttt{null});
  \item include comprehensive evidence excerpts across notes with exact \texttt{note\_id}, \texttt{line\_no}, and verbatim \texttt{span}.
\end{itemize}

\section{System design}

The system is implemented as a two-stage LLM extraction pipeline with strict post-validation and caching.

\subsection{Stage A: Note-level extraction}
For each note:
\begin{itemize}[leftmargin=1.2em]
  \item Load the note and assign 1-indexed line numbers.
  \item Prompt an OpenAI-compatible LLM to extract only taxonomy-valid conditions mentioned in that note.
  \item Enforce evidence integrity: each evidence span must be copied verbatim from the referenced line; if not, fall back to the full line.
  \item Validate extracted objects against a schema (Pydantic) and taxonomy keys.
\end{itemize}

\subsection{Stage B: Patient-level consolidation}
We consolidate candidates across notes to produce the final patient condition inventory:
\begin{itemize}[leftmargin=1.2em]
  \item De-duplicate near-identical conditions within the same taxonomy slot using lightweight fuzzy matching.
  \item Set \texttt{status} as of the latest note where the condition appears (note order provided to the consolidator).
  \item Set \texttt{onset} using the task’s priority rules (model instructed).
  \item Aggregate evidence across all notes where mentioned.
\end{itemize}

\subsection{Caching, cost, and determinism}
LLM calls are cached on disk keyed by a hash of prompt contents; re-runs on the same inputs avoid duplicate token consumption. Default temperature is 0.0 to reduce variance.

\section{Pipeline diagram (figure)}

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=10mm,
  box/.style={draw, rounded corners, align=center, minimum width=42mm, minimum height=10mm},
  arr/.style={-{Latex[length=3mm]}, thick}
]
  \node[box] (input) {Patient folder\\\texttt{patient\_XX/}\\\texttt{text\_0..N.md}};
  \node[box, right=14mm of input] (note) {Stage A\\Note extraction\\(line-numbered)};
  \node[box, right=14mm of note] (cand) {Candidate conditions\\per note\\(taxonomy-valid)};
  \node[box, right=14mm of cand] (cons) {Stage B\\Patient consolidation\\(dedupe, onset, status)};
  \node[box, right=14mm of cons] (out) {Output\\\texttt{output/patient\_XX.json}};

  \draw[arr] (input) -- (note);
  \draw[arr] (note) -- (cand);
  \draw[arr] (cand) -- (cons);
  \draw[arr] (cons) -- (out);
\end{tikzpicture}
\caption{End-to-end extraction pipeline.}
\end{figure}

\section{Evidence completeness}
\label{sec:evidence}

The specification requires evidence to be comprehensive: include excerpts from every note where a condition is mentioned. The current system enforces this primarily by (i) supplying per-note candidates from all notes and (ii) explicitly instructing the patient-level consolidator to include evidence from every note where mentioned, using only provided spans.

\subsection{Recommended hardening}
To reduce dependence on LLM compliance, add a deterministic post-check:
\begin{itemize}[leftmargin=1.2em]
  \item For each predicted condition, scan all note lines for normalized mention strings (and common abbreviations/ICD tokens observed in training data).
  \item If a mention exists in a note without evidence, append a new evidence entry referencing the matching line.
\end{itemize}
This improves evidence recall without extra LLM calls.

\section{Experiments and results (template)}

The repository includes a lightweight evaluation helper for iteration on labeled training patients:
\begin{quote}
\texttt{python -m clinical\_nlp\_assignment.train --data-dir ./clinical\_nlp\_assignment/train --taxonomy-path ./clinical\_nlp\_assignment/taxonomy.json}
\end{quote}

We report a practical dev metric using exact taxonomy matching and fuzzy matching over normalized condition names. Fill the table below after running:

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
Split & Precision & Recall & F1 \\
\midrule
Train (macro) & \_\_ & \_\_ & \_\_ \\
\bottomrule
\end{tabular}
\caption{Development metrics (for iteration; official evaluation may differ).}
\end{table}

\section{Reproducibility and running}

\subsection{Dependencies}
\begin{quote}
\texttt{pip install -r requirements.txt}
\end{quote}

\subsection{Environment variables}
The evaluator provides these at runtime:
\begin{itemize}[leftmargin=1.2em]
  \item \texttt{OPENAI\_BASE\_URL}
  \item \texttt{OPENAI\_API\_KEY}
  \item \texttt{OPENAI\_MODEL}
\end{itemize}

\subsection{Inference command}
\begin{quote}
\texttt{python main.py --data-dir ./clinical\_nlp\_assignment/dev --patient-list ./patients.json --output-dir ./output}
\end{quote}

\subsection{Output validation}
\begin{quote}
\texttt{python -m clinical\_nlp\_assignment.validate\_outputs --output-dir ./output --taxonomy-path ./clinical\_nlp\_assignment/taxonomy.json}
\end{quote}

\section{What worked, what didn’t}

\subsection{What worked}
\begin{itemize}[leftmargin=1.2em]
  \item Two-stage extraction helps recall in long documents and encourages comprehensive evidence.
  \item Strict schema + taxonomy validation prevents invalid outputs.
  \item Disk caching significantly reduces repeat cost/time.
\end{itemize}

\subsection{Risks / limitations}
\begin{itemize}[leftmargin=1.2em]
  \item Some OpenAI-compatible providers may not support \texttt{response\_format=\{"type":"json\_object"\}}; if so, the client should fall back to JSON-only prompting + parsing.
  \item Evidence completeness can still depend on model behavior unless hardened as described in Section~\ref{sec:evidence}.
\end{itemize}

\end{document}

